Sequence modeling enables direct inference of germline haplotypes from short read sequencing data

### 2022-08-09 manuscript/manuscript.tex

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting low-complexity regions. We additionally included `true negative' regions where no known variant was present.  For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.  We examine three different sizes of training data sets with 5.1M, 10.2M and 20.4M examples.

@@

@@

Models were trained for 30 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$ and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).

For a given haplotype we use a standard cross-entropy loss function to compare the prediction to the target sequence. However, two haplotypes exist for each region and they have no inherent order. For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn predict which because the assigment is random. To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.

@@

The above variant detection procedure has two shortcomings. First, if every potential variant is emitted, then specificity is poor. Second, the model does not innately produce well calibrated variant quality scores. To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).  The resulting score is used as a final variant quality score suitable for tuning the tradeoff between sensitivity and specificity.

@@

We explore the effects of model size and training data size on validation loss and variant detection accuracy. We evaluate three model sizes and two sets of training data.

@@

@@

>>>>

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions containing a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting low-complexity regions. We additionally included `true negative' regions where no known variant was present. In regions containing more than 100 overlapping reads, the reads were repeatedly downsampled to generate multiple training regions. For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation. Approximately 1.3M regions were obtained from each of the 17 samples, generating a full training data size of 24.3M regions. In addition to the "Full" training set we also examine a half-sized training data set of 12.1M regions, and quarter-sized set of 6.0M regions. The number of regions generated for each variant type is shown in Table \ref{table:regioncounts}.

@@

\begin{center}
	\begin{tabular}{ lc }
	 \textbf{Region type} & \textbf{Number of regions (per sample)} \\
	 No variant & 200,000 \\
	 SNV & 200,000 \\
	 Insertions & 200,000 \\
	 Deletions & 200,000 \\
	 Insertion/Deletion + SNV & 50,000 \\
	 Insertion + Deletion & 30,000 \\
	 SNV (low complexity) & 80,000 \\
	 Insertion (low complexity) & 80,000 \\
	 Deletion (low complexity) & 80,000 \\  
	 Insertion + Deletion (low complexity) & 20,000 \\
	 Insertion/Deletion + SNV (low complexity) & 80,000 \\
	\end{tabular}
	\label{table:regioncounts}
\end{center}

@@

Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e^{-5}$ and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).

For a given haplotype we use a modified cross-entropy loss function to compare the predictions to the target sequence. However, two haplotypes exist for each region and they have no inherent order. For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn predict which because the assigment is random. To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.

@@

The above variant detection procedure has two shortcomings. First, if every potential variant is emitted, then precision is poor. Second, the model does not innately produce well calibrated variant quality scores. To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table \ref{table:features} for list of features).  The resulting score is used as a final variant quality score suitable for tuning the tradeoff between sensitivity and specificity.

@@

We explore the effects of model size and training data size on variant detection accuracy. Throughout, we calculate sensitivity and precision on a small set of evaluation regions from chromosomes 21 and 22 that were not used for training. Importantly, for performance reasons these sensitivity and precision calculations do \textit{not} include the full variant detection procedure with multiple overlapping windows and the post-calling classifier, and instead reflect results from a single inference procedure on the target window (accuracy of the full detection procedure is explored below).  We evaluate three model sizes and three sets of training data (Full, Half, and Quarter-sized, see section Training Data above for details).

@@

During training, sensitivity and precision as measured on the holdout regions from chromosomes 21 and 22 increase for all variant classes, with returns dwindling after 30-40 epochs of training. Little difference was observed between the "full" and "half" sized training data sets, but both sensitivity and precision were lower for the quarter-sied training set. These results indicate that additional training examples beyond the "half" size ($12e^6$ example regions) offer little improvement in accuracy, and suggest that performance becomes limited by other factors, such as model capacity, inaccuracies in the training data set, or a non-optimal loss function.

\begin{figure}[htp]
	\includegraphics[scale=0.5]{./trainsize_effect.png}
	\caption{ Sensitivity and precision curves for the 10M model during training for different training data sizes. Dashed lines are raw values, solid lines indicate rolling average with window size of 5. }
	\label{fig:trainsize}
\end{figure}

@@

Limitations
 - Low SNV precision
 - not using strand bias?
 -
